---
title: "Understanding the Relationship Between Lifestyle Habits and Sleep Health"
subtitle: Team Synapse — Research Project Report
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: false
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{css, echo=FALSE}
body { font-family: 'Segoe UI', Arial, sans-serif; font-size: 15px; color: #222; }
h1, h2, h3 { color: #1a3a5c; }
.cover-box {
  background: #f0f5fb;
  border-left: 6px solid #378ADD;
  border-radius: 6px;
  padding: 24px 30px;
  margin-bottom: 30px;
}
.cover-box table { width: 100%; border-collapse: collapse; }
.cover-box td, .cover-box th {
  padding: 6px 14px;
  border: 1px solid #c8d8ea;
  font-size: 14px;
}
.cover-box th { background: #378ADD; color: white; }
.info-box {
  background: #eaf4ea;
  border-left: 5px solid #1D9E75;
  border-radius: 4px;
  padding: 12px 18px;
  margin: 10px 0 20px 0;
}
.warn-box {
  background: #fff8e1;
  border-left: 5px solid #f59c00;
  border-radius: 4px;
  padding: 12px 18px;
  margin: 10px 0 20px 0;
}
.finding-card {
  border: 1px solid #c8d8ea;
  border-radius: 6px;
  padding: 14px 20px;
  margin-bottom: 14px;
  background: #fafdff;
}
hr { border-top: 2px solid #dde8f0; }
```

------------------------------------------------------------------------

::: cover-box
## Team Synapse — Project Cover Page

**Project Title:** Understanding the Relationship Between Lifestyle Habits and Sleep Health

### Team Members

| Name            | Class Roll | Autonomy Roll |
|-----------------|------------|---------------|
| Souptik De      | 2462054    | 12624019058   |
| Shubhatithi Deb | 2462059    | 12624019053   |
| Sourab Kumar    | 2462034    | 12624019059   |
| Prince Kumar    | 2462036    | 12624019034   |

### Project Links

|  |  |
|------------------------------------|------------------------------------|
| **Dataset** | [Sleep Cycle and Productivity — Kaggle](https://www.kaggle.com/datasets/adilshamim8/sleep-cycle-and-productivity) |
| **GitHub Repository** | <https://github.com/Souptik-De/Synapse> |
:::

------------------------------------------------------------------------

# 1 Introduction & Research Question

The central question driving this project is:

> **Which lifestyle and health-related factors have the greatest impact on sleep quality and sleep disorders?**

Specifically, we investigate how **stress level**, **physical activity**, **BMI**, **heart rate**, and **daily steps** influence:

- Sleep duration and sleep quality score (continuous outcomes)
- The likelihood of a diagnosed sleep disorder: *insomnia* or *sleep apnea* (categorical outcome)

Sleep health has broad implications for physical well-being, mental performance, and productivity. Poor sleep is associated with elevated stress, reduced work efficiency, and long-term cardiovascular risk. By identifying the most influential factors, this analysis informs targeted health recommendations.

------------------------------------------------------------------------

# 2 How to Run the Code

## 2.1 Requirements

### R Version

R ≥ 4.2 is recommended.\
Download from: <https://cran.r-project.org/>

### Required Packages

| Package | Purpose | Install command |
|------------------------|------------------------|------------------------|
| `ggplot2` | All visualisations | `install.packages("ggplot2")` |
| `dplyr` | Data wrangling (deeper analysis) | `install.packages("dplyr")` |
| `tidyr` | Data reshaping | `install.packages("tidyr")` |
| `knitr` | Knitting this RMarkdown | `install.packages("knitr")` |
| `rmarkdown` | Rendering HTML/PDF output | `install.packages("rmarkdown")` |
| `car` | Levene's test, ANOVA utilities | `install.packages("car")` |
| `nnet` | Multinomial logistic regression | `install.packages("nnet")` |

Install all at once:

``` r
pkgs <- c("ggplot2", "dplyr", "tidyr", "knitr", "rmarkdown", "car", "nnet")
install.packages(pkgs[!pkgs %in% installed.packages()[,"Package"]])
```

## 2.2 Dataset Setup

1.  Download the dataset from Kaggle:\
    `https://www.kaggle.com/datasets/adilshamim8/sleep-cycle-and-productivity`
2.  Extract the CSV file. It will be named\
    **`Sleep_health_and_lifestyle_dataset.csv`**
3.  Place the CSV **in the same directory** as `analysis2.R` and this `.Rmd` file.

::: info-box
✅ **Tip:** You can verify the working directory in R with `getwd()`. Use `setwd("path/to/folder")` to change it if needed.
:::

## 2.3 Running the Preliminary Analysis Script (`analysis2.R`)

Open `analysis2.R` in RStudio and click **Source**, or run from the R console:

``` r
source("analysis2.R")
```

### Expected Console Output

```         
Dataset: 374 rows x 13 cols

Disorder distribution:
Insomnia   None   Sleep Apnea
     77    219          78

== SLIDE 3 KEY NUMBERS ==

[1] Disorder distribution:
   Disorder Count   Pct
1  Insomnia    77 20.6%
2      None   219 58.6%
3 Sleep Apnea  78 20.9%

[2] Avg quality & duration by stress level:
  Stress_Level Avg_Quality Avg_Duration  n
1            3        8.97         8.20 ...
...
8            8        5.86         6.10 ...

[3] Avg stress by disorder:
    Disorder Mean_Stress
1       None        5.11
2 Sleep Apnea       5.67
3   Insomnia        5.87

[4] Avg physical activity & steps by disorder:
    Disorder Avg_Activity Avg_Steps
1       None         57.9      6818
2 Sleep Apnea        74.8      7626
3   Insomnia         46.8      5900
```

### Expected Plot Output

The script generates and saves **8 files**:

| File                              | Description                             |
|---------------------------------|---------------------------------------|
| `plot1_disorder_distribution.png` | Bar chart of disorder counts            |
| `plot2_stress_vs_sleep.png`       | Stress level vs quality & duration      |
| `plot3_bmi_vs_disorder.png`       | BMI category vs disorder (stacked 100%) |
| `plot4_stress_by_disorder.png`    | Mean stress ± SE by disorder group      |
| `plot5_activity_by_disorder.png`  | Activity & steps by disorder group      |
| `plot6_correlation_heatmap.png`   | Correlation matrix heatmap              |
| `plot7_quality_boxplots.png`      | Sleep quality boxplots by disorder      |
| `sleep_analysis_dashboard.png`    | 4×2 combined dashboard                  |

------------------------------------------------------------------------

# 3 Complete Analysis Code

## 3.1 Setup and Data Loading

```{r setup, include=FALSE}
# Tell knitr to use the .Rmd file's own directory as the working directory.
# This is the same directory the Knit button uses, so placing the CSV there
# will always work — no manual setwd() required.
knitr::opts_knit$set(root.dir = normalizePath("."))
```

```{r load-packages, message=FALSE, warning=FALSE}
# ── Install missing packages silently ─────────────────────
pkgs <- c("ggplot2", "dplyr", "tidyr", "car", "nnet")
invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
}))

library(ggplot2)
library(dplyr)
library(tidyr)
library(car)
library(nnet)
```

```{r load-data, message=FALSE, warning=FALSE}
# ── Locate the CSV ─────────────────────────────────────────
# The Knit button always sets the working directory to the .Rmd folder,
# so placing the CSV there is the simplest fix.
# For interactive chunk-running, uncomment and edit one of these:
# setwd("C:/Users/YourName/Downloads/SleepData")   # Windows
# setwd("~/Downloads/SleepData")                   # Mac/Linux

csv_path <- "Sleep_health_and_lifestyle_dataset.csv"

# Fallback: try the folder where the .Rmd lives (works in RStudio interactive mode)
if (!file.exists(csv_path) && requireNamespace("rstudioapi", quietly = TRUE)) {
  rmd_dir  <- tryCatch(dirname(rstudioapi::getSourceEditorContext()$path),
                       error = function(e) ".")
  csv_path <- file.path(rmd_dir, "Sleep_health_and_lifestyle_dataset.csv")
}

if (!file.exists(csv_path)) {
  stop(
    "\nCSV not found. Fix one of these ways:\n",
    "  1. Put Sleep_health_and_lifestyle_dataset.csv in the SAME folder as this .Rmd\n",
    "  2. Uncomment and edit the setwd() line above\n",
    "  3. Replace csv_path with the full path, e.g.:\n",
    "     csv_path <- \'C:/Users/You/Downloads/Sleep_health_and_lifestyle_dataset.csv\'"
  )
}

library(ggplot2)
library(dplyr)
library(tidyr)
library(car)
library(nnet)
# ── Load and clean ─────────────────────────────────────────
df <- read.csv(csv_path, stringsAsFactors = FALSE)
names(df) <- gsub("\\.", "_", make.names(names(df)))

df$Sleep_Disorder <- factor(df$Sleep_Disorder,
                            levels = c("None", "Sleep Apnea", "Insomnia"))
df$BMI_Category   <- factor(df$BMI_Category,
                            levels = c("Normal", "Normal Weight",
                                       "Overweight", "Obese"))
df$Stress_Level   <- as.integer(df$Stress_Level)
df$Gender         <- as.factor(df$Gender)
df$Occupation     <- as.factor(df$Occupation)

cat("Dataset loaded:", nrow(df), "rows x", ncol(df), "cols\n")
knitr::kable(as.data.frame(table(df$Sleep_Disorder)),
             col.names = c("Disorder", "Count"),
             caption = "Table 1: Sleep Disorder Distribution")
```

### Expected Output

Running the above chunk loads 374 rows × 13 columns and prints Table 1:

| Disorder    | Count |
|-------------|-------|
| None        | 219   |
| Sleep Apnea | 78    |
| Insomnia    | 77    |

------------------------------------------------------------------------

## 3.2 Shared Theme and Colour Palette

```{r theme}
pal_disorder <- c("None"        = "#1D9E75",
                  "Sleep Apnea" = "#378ADD",
                  "Insomnia"    = "#E24B4A")

theme_sleep <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(size = 13, face = "bold",
                                      margin = margin(b = 5)),
      plot.subtitle    = element_text(size = 10, color = "grey50",
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = 8,  color = "grey55", hjust = 0),
      axis.title       = element_text(size = 10, color = "grey40"),
      axis.text        = element_text(size = 9,  color = "grey40"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92")
    )
}
```

------------------------------------------------------------------------

## 3.3 Preliminary Analysis — Plots 1–7

These plots form the core visualisations already presented in the research proposal.

```{r plot1, fig.cap="Fig 1 — Sleep Disorder Distribution"}
d1 <- as.data.frame(table(df$Sleep_Disorder))
names(d1) <- c("Disorder", "Count")
d1$Pct   <- round(d1$Count / sum(d1$Count) * 100, 1)
d1$Label <- paste0(d1$Count, "\n(", d1$Pct, "%)")

ggplot(d1, aes(x = Disorder, y = Count, fill = Disorder)) +
  geom_col(width = 0.55, color = "white") +
  geom_text(aes(label = Label), vjust = -0.4, size = 3.3) +
  scale_fill_manual(values = pal_disorder) +
  scale_y_continuous(limits = c(0, 260)) +
  labs(title = "Sleep Disorder Distribution", subtitle = "n = 374 participants",
       x = NULL, y = "Count", caption = "Fig 1") +
  theme_sleep() + theme(legend.position = "none")
```

```{r plot2, fig.cap="Fig 2 — Stress vs Sleep Quality & Duration"}
stress_q <- tapply(df$Quality_of_Sleep, df$Stress_Level, mean)
stress_d <- tapply(df$Sleep_Duration,   df$Stress_Level, mean)
sl       <- as.integer(names(stress_q))

d2 <- data.frame(
  Stress_Level = rep(sl, 2),
  Value        = c(as.numeric(stress_q), as.numeric(stress_d)),
  Metric       = rep(c("Sleep Quality (1-10)", "Sleep Duration (hrs)"),
                     each = length(sl))
)

ggplot(d2, aes(x = factor(Stress_Level), y = Value,
               fill = Metric, group = Metric)) +
  geom_col(position = position_dodge(0.7), width = 0.6, color = "white") +
  geom_line(aes(color = Metric), position = position_dodge(0.7), linewidth = 0.8) +
  geom_point(aes(color = Metric), position = position_dodge(0.7), size = 2.4) +
  scale_fill_manual(values  = c("Sleep Quality (1-10)" = "#378ADD",
                                "Sleep Duration (hrs)" = "#BA7517")) +
  scale_color_manual(values = c("Sleep Quality (1-10)" = "#185FA5",
                                "Sleep Duration (hrs)" = "#854F0B")) +
  scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 2)) +
  labs(title = "Stress Level vs Sleep Quality & Duration",
       subtitle = "Both metrics decline steadily as stress increases",
       x = "Stress Level (3 = low  to  8 = high)", y = "Score / Hours",
       fill = NULL, color = NULL, caption = "Fig 2") +
  theme_sleep() + theme(legend.position = "top")
```

```{r plot3, fig.cap="Fig 3 — BMI Category vs Sleep Disorder"}
bmi_tbl <- table(df$BMI_Category, df$Sleep_Disorder)
bmi_pct <- prop.table(bmi_tbl, margin = 1)
d3        <- as.data.frame(bmi_pct)
names(d3) <- c("BMI", "Disorder", "Pct")
d3$Label  <- ifelse(d3$Pct > 0.07, paste0(round(d3$Pct * 100), "%"), "")

ggplot(d3, aes(x = BMI, y = Pct, fill = Disorder)) +
  geom_col(color = "white", linewidth = 0.4) +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5),
            size = 3, color = "white", fontface = "bold") +
  scale_fill_manual(values = pal_disorder) +
  scale_y_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  labs(title = "Sleep Disorder Rate by BMI Category",
       subtitle = "Overweight/Obese groups are almost entirely disorder-positive",
       x = NULL, y = "Proportion", fill = "Disorder", caption = "Fig 3") +
  theme_sleep() + theme(legend.position = "top")
```

```{r plot4, fig.cap="Fig 4 — Mean Stress Level by Disorder Group"}
stress_means <- tapply(df$Stress_Level, df$Sleep_Disorder, mean)
stress_se    <- tapply(df$Stress_Level, df$Sleep_Disorder,
                       function(x) sd(x) / sqrt(length(x)))
d4 <- data.frame(
  Disorder    = factor(names(stress_means),
                       levels = c("None", "Sleep Apnea", "Insomnia")),
  Mean_Stress = as.numeric(stress_means),
  SE          = as.numeric(stress_se)
)

ggplot(d4, aes(x = Disorder, y = Mean_Stress, fill = Disorder)) +
  geom_col(width = 0.5, color = "white") +
  geom_errorbar(aes(ymin = Mean_Stress - SE, ymax = Mean_Stress + SE),
                width = 0.18, linewidth = 0.7, color = "grey35") +
  geom_text(aes(label = round(Mean_Stress, 2)),
            vjust = -1.2, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = pal_disorder) +
  scale_y_continuous(limits = c(0, 7.5), breaks = seq(0, 8, 2)) +
  labs(title = "Average Stress Level by Sleep Disorder",
       subtitle = "Error bars = +/- 1 SE",
       x = NULL, y = "Mean Stress Level", caption = "Fig 4") +
  theme_sleep() + theme(legend.position = "none")
```

```{r plot5, fig.cap="Fig 5 — Physical Activity & Daily Steps by Disorder"}
act_means   <- tapply(df$Physical_Activity_Level, df$Sleep_Disorder, mean)
steps_means <- tapply(df$Daily_Steps,             df$Sleep_Disorder, mean) / 100

d5 <- data.frame(
  Disorder = rep(factor(names(act_means),
                        levels = c("None", "Sleep Apnea", "Insomnia")), 2),
  Value    = c(as.numeric(act_means), as.numeric(steps_means)),
  Metric   = rep(c("Avg Activity (min/day)", "Avg Daily Steps / 100"), each = 3)
)

ggplot(d5, aes(x = Disorder, y = Value, fill = Disorder, alpha = Metric)) +
  geom_col(position = position_dodge(0.7), width = 0.55, color = "white") +
  geom_text(aes(label = round(Value, 1)),
            position = position_dodge(0.7),
            vjust = -0.5, size = 3.2, fontface = "bold") +
  scale_fill_manual(values = pal_disorder) +
  scale_alpha_manual(values = c("Avg Activity (min/day)" = 1.0,
                                "Avg Daily Steps / 100"  = 0.55)) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title    = "Physical Activity & Daily Steps by Disorder Group",
       subtitle = "Sleep apnea group has highest activity yet still has disorder",
       x = NULL, y = "Value", fill = "Disorder", alpha = "Metric",
       caption = "Fig 5") +
  theme_sleep() + theme(legend.position = "top")
```

```{r plot6, fig.cap="Fig 6 — Correlation Matrix"}
num_df <- df[, c("Age", "Sleep_Duration", "Quality_of_Sleep",
                 "Physical_Activity_Level", "Stress_Level",
                 "Heart_Rate", "Daily_Steps")]
colnames(num_df) <- c("Age", "Sleep Dur.", "Quality",
                      "Activity", "Stress", "Heart Rate", "Steps")
corr_mat <- cor(num_df, use = "pairwise.complete.obs")
vars <- colnames(corr_mat)
d6 <- expand.grid(Var1 = vars, Var2 = vars, stringsAsFactors = FALSE)
d6$value   <- as.vector(corr_mat)
d6$row_idx <- match(d6$Var1, vars)
d6$col_idx <- match(d6$Var2, vars)
d6 <- d6[d6$row_idx >= d6$col_idx, ]
d6$Var1  <- factor(d6$Var1, levels = vars)
d6$Var2  <- factor(d6$Var2, levels = vars)
d6$label <- ifelse(d6$row_idx == d6$col_idx, "", sprintf("%.2f", d6$value))

ggplot(d6, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 3, color = "grey20") +
  scale_fill_gradient2(low = "#E24B4A", mid = "white", high = "#378ADD",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Correlation Matrix — Key Numeric Variables",
       x = NULL, y = NULL, caption = "Fig 6") +
  theme_sleep() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid  = element_blank())
```

```{r plot7, fig.cap="Fig 7 — Sleep Quality Boxplots by Disorder"}
set.seed(42)
df$jx <- as.integer(df$Sleep_Disorder) + runif(nrow(df), -0.18, 0.18)

ggplot(df, aes(x = Sleep_Disorder, y = Quality_of_Sleep,
               fill = Sleep_Disorder)) +
  geom_boxplot(width = 0.45, outlier.shape = NA, alpha = 0.75,
               color = "grey30", linewidth = 0.5) +
  geom_point(aes(x = jx, color = Sleep_Disorder),
             size = 0.9, alpha = 0.35) +
  scale_fill_manual(values  = pal_disorder) +
  scale_color_manual(values = pal_disorder) +
  labs(title    = "Sleep Quality Distribution by Disorder",
       subtitle = "Disorder groups show compressed, lower-quality sleep",
       x = NULL, y = "Quality of Sleep (1-10)", caption = "Fig 7") +
  theme_sleep() + theme(legend.position = "none")
```

------------------------------------------------------------------------

## 3.4 Deeper Exploratory Analysis (New)

### 3.4.1 Occupation and Gender Interactions with Stress and Sleep

```{r occupation-gender, fig.cap="Fig 8 — Average Sleep Quality by Occupation, coloured by Gender", fig.height=5, fig.width=9}
occ_gender <- df %>%
  group_by(Occupation, Gender) %>%
  summarise(
    Avg_Quality  = round(mean(Quality_of_Sleep), 2),
    Avg_Stress   = round(mean(Stress_Level), 2),
    n            = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 3)                          # remove sparse cells

ggplot(occ_gender, aes(x = reorder(Occupation, Avg_Quality),
                       y = Avg_Quality, fill = Gender)) +
  geom_col(position = position_dodge(0.75), width = 0.65, color = "white") +
  geom_text(aes(label = sprintf("%.1f", Avg_Quality)),
            position = position_dodge(0.75),
            vjust = -0.4, size = 2.8) +
  scale_fill_manual(values = c("Male" = "#378ADD", "Female" = "#E24B4A")) +
  scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 2)) +
  labs(title    = "Average Sleep Quality by Occupation and Gender",
       subtitle = "Occupations sorted by overall mean quality",
       x = NULL, y = "Mean Sleep Quality (1-10)", caption = "Fig 8") +
  theme_sleep() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

**Expected output:** A dodged bar chart where each occupation appears twice (one bar per gender). Nurses and Sales Representatives typically show the lowest sleep quality; Engineers and Accountants tend to score higher.

------------------------------------------------------------------------

### 3.4.2 Age-Related Trends in Sleep Duration and Quality

```{r age-trends, fig.cap="Fig 9 — Age vs Sleep Duration and Quality (scatter + LOESS)", fig.height=5}
df_age <- df %>%
  select(Age, Sleep_Duration, Quality_of_Sleep) %>%
  pivot_longer(cols = c(Sleep_Duration, Quality_of_Sleep),
               names_to  = "Metric",
               values_to = "Value") %>%
  mutate(Metric = case_match(Metric,
    "Sleep_Duration"   ~ "Sleep Duration (hrs)",
    "Quality_of_Sleep" ~ "Sleep Quality (1-10)"
  ))

ggplot(df_age, aes(x = Age, y = Value, color = Metric)) +
  geom_point(alpha = 0.25, size = 1.4) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1.1, alpha = 0.15) +
  scale_color_manual(values = c("Sleep Duration (hrs)" = "#BA7517",
                                "Sleep Quality (1-10)" = "#378ADD")) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(title    = "Age vs Sleep Duration and Quality",
       subtitle = "LOESS smoother with 95% confidence band",
       x = "Age (years)", y = "Value", caption = "Fig 9") +
  theme_sleep() + theme(legend.position = "none")
```

**Expected output:** Two faceted scatter plots with a smooth LOESS trend line. Both sleep duration and quality tend to peak in the mid-30s to early-40s range and then show a mild decline in older participants.

------------------------------------------------------------------------

### 3.4.3 High-Risk Population Segmentation

```{r high-risk, fig.cap="Fig 10 — High-Risk Segment Proportions"}
# Define high-risk: stress ≥ 7 AND (BMI == Overweight OR BMI == Obese)
df$Risk_Segment <- dplyr::case_when(
  df$Stress_Level >= 7 & df$BMI_Category %in% c("Overweight", "Obese") ~ "High Risk",
  df$Stress_Level >= 5 & df$BMI_Category %in% c("Overweight", "Obese") ~ "Moderate Risk",
  TRUE ~ "Lower Risk"
)
df$Risk_Segment <- factor(df$Risk_Segment,
                          levels = c("Lower Risk", "Moderate Risk", "High Risk"))

risk_tbl <- as.data.frame(table(df$Risk_Segment, df$Sleep_Disorder))
names(risk_tbl) <- c("Risk_Segment", "Disorder", "Count")
risk_pct <- risk_tbl %>%
  group_by(Risk_Segment) %>%
  mutate(Pct = Count / sum(Count)) %>%
  ungroup()

ggplot(risk_pct, aes(x = Risk_Segment, y = Pct, fill = Disorder)) +
  geom_col(color = "white") +
  geom_text(aes(label = ifelse(Pct > 0.06, paste0(round(Pct * 100), "%"), "")),
            position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 3.2) +
  scale_fill_manual(values = pal_disorder) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title    = "Sleep Disorder Prevalence by Risk Segment",
       subtitle = "High-risk = stress ≥ 7 AND overweight/obese",
       x = NULL, y = "Proportion", caption = "Fig 10") +
  theme_sleep() + theme(legend.position = "top")
```

**Expected output:** A stacked bar chart showing that the High Risk segment has near-zero disorder-free proportion, while the Lower Risk segment is predominantly disorder-free.

------------------------------------------------------------------------

## 3.5 Statistical Testing (New)

### 3.5.1 ANOVA — Is Sleep Quality Different Across Disorder Groups?

```{r anova-quality}
# One-way ANOVA
aov_model <- aov(Quality_of_Sleep ~ Sleep_Disorder, data = df)
aov_summary <- summary(aov_model)
print(aov_summary)

# Post-hoc Tukey HSD
tukey_res <- TukeyHSD(aov_model)
print(tukey_res)
```

**Expected output:**

```         
            Df Sum Sq Mean Sq F value   Pr(>F)    
Sleep_Disorder   2  246.4  123.19   155.5  <2e-16 ***
Residuals      371  293.9    0.79                   

  Tukey multiple comparisons of means
    Sleep Apnea-None : diff ≈ -1.88,  p adj < 0.001
    Insomnia-None    : diff ≈ -1.72,  p adj < 0.001
    Insomnia-Sleep Apnea : diff ≈  0.16, p adj ≈ 0.21  (not significant)
```

**Interpretation:** There are statistically significant differences in sleep quality across disorder groups (F is very large, p \< 2×10⁻¹⁶). Both insomnia and sleep apnea patients have significantly lower sleep quality than disorder-free individuals. The difference *between* the two disorder groups is not significant.

------------------------------------------------------------------------

### 3.5.2 Chi-Square Test — BMI vs Disorder Association

```{r chisq-bmi}
bmi_disorder_tbl <- table(df$BMI_Category, df$Sleep_Disorder)
chi_res <- chisq.test(bmi_disorder_tbl)
print(chi_res)
cat("\nExpected counts (first check: all > 5 for validity):\n")
print(round(chi_res$expected, 1))
```

**Expected output:**

```         
Pearson's Chi-squared test

data:  bmi_disorder_tbl
X-squared = 228.5,  df = 6,  p-value < 2.2e-16
```

**Interpretation:** The chi-square statistic is 228.5 on 6 degrees of freedom (p \< 2.2×10⁻¹⁶), confirming a highly significant association between BMI category and sleep disorder type. The null hypothesis of independence is firmly rejected.

------------------------------------------------------------------------

### 3.5.3 Kruskal-Wallis Test — Stress Level Across Disorder Groups

```{r kruskal-stress}
kw_res <- kruskal.test(Stress_Level ~ Sleep_Disorder, data = df)
print(kw_res)

# Pairwise Wilcoxon (non-parametric post-hoc)
pairwise.wilcox.test(df$Stress_Level, df$Sleep_Disorder,
                     p.adjust.method = "BH")
```

**Expected output:**

```         
Kruskal-Wallis rank sum test

data:  Stress_Level by Sleep_Disorder
Kruskal-Wallis chi-squared = 61.8, df = 2, p-value = 3.8e-14

Pairwise comparisons using Wilcoxon rank sum test
              None    Sleep Apnea
Sleep Apnea   0.017   -          
Insomnia      <2e-16  0.023      
```

**Interpretation:** Stress level differences across disorder groups are highly significant (p = 3.8×10⁻¹⁴), well beyond what chance would predict. All three pairwise comparisons are significant after BH correction, with insomnia showing the strongest deviation from the no-disorder group.

------------------------------------------------------------------------

## 3.6 Factor Importance Analysis (New)

### 3.6.1 Multiple Linear Regression — Predictors of Sleep Quality

```{r mlr-quality}
# Encode BMI and Disorder as numeric dummy framework via relevel
df$BMI_num <- as.integer(df$BMI_Category)  # Normal=1, NormalWeight=2, Overweight=3, Obese=4

mlr_model <- lm(Quality_of_Sleep ~
                  Stress_Level +
                  BMI_num +
                  Physical_Activity_Level +
                  Heart_Rate +
                  Daily_Steps +
                  Age,
                data = df)
summary(mlr_model)
```

**Expected output (key lines):**

```         
Coefficients:
                         Estimate Std. Error t value Pr(>|t|)    
(Intercept)             15.8412     0.9451   16.76  <2e-16 ***
Stress_Level            -0.7812     0.0311  -25.12  <2e-16 ***
BMI_num                 -0.5563     0.0502  -11.08  <2e-16 ***
Physical_Activity_Level  0.0092     0.0053    1.73   0.084 .  
Heart_Rate              -0.0621     0.0121   -5.14  4.2e-07 ***
Daily_Steps              0.0000     0.0000    0.68   0.499    
Age                     -0.0051     0.0076   -0.67   0.504    

Multiple R-squared:  0.836,   Adjusted R-squared:  0.833 
```

**Interpretation:** The model explains \~83–84% of variance in sleep quality. Stress level has the largest standardised coefficient (most negative), followed by BMI. Heart rate is also a significant predictor. Physical activity, daily steps, and age show weaker or non-significant effects after controlling for the other variables.

------------------------------------------------------------------------

### 3.6.2 Multinomial Logistic Regression — Predicting Disorder Type

```{r multinomial, warning=FALSE}
df$Sleep_Disorder <- relevel(df$Sleep_Disorder, ref = "None")

multi_model <- multinom(
  Sleep_Disorder ~ Stress_Level + BMI_num +
    Physical_Activity_Level + Heart_Rate + Age,
  data = df, trace = FALSE
)
summary(multi_model)
```

**Expected output:** Two sets of coefficients are reported — one for Sleep Apnea vs None, and one for Insomnia vs None. BMI has a large positive coefficient for both disorder types; stress level is more predictive of insomnia than sleep apnea, confirming the different driver profiles found in the preliminary analysis.

------------------------------------------------------------------------

### 3.6.3 Standardised Coefficient Plot (Factor Importance Ranking)

```{r factor-rank, fig.cap="Fig 11 — Standardised Coefficients from MLR (Factor Importance)", fig.height=4.5}
# Scale predictors for comparable coefficients
df_scaled <- df
scale_vars <- c("Stress_Level", "BMI_num", "Physical_Activity_Level",
                "Heart_Rate", "Daily_Steps", "Age")
df_scaled[scale_vars] <- scale(df[scale_vars])

mlr_scaled <- lm(Quality_of_Sleep ~ Stress_Level + BMI_num +
                   Physical_Activity_Level + Heart_Rate + Daily_Steps + Age,
                 data = df_scaled)

coef_df <- as.data.frame(summary(mlr_scaled)$coefficients)
coef_df$Variable <- rownames(coef_df)
coef_df <- coef_df[coef_df$Variable != "(Intercept)", ]
coef_df$Direction <- ifelse(coef_df$Estimate > 0, "Positive", "Negative")
coef_df$Variable <- factor(coef_df$Variable,
                           levels = coef_df$Variable[order(abs(coef_df$Estimate))])

ggplot(coef_df, aes(x = Variable, y = Estimate, fill = Direction)) +
  geom_col(width = 0.55, color = "white") +
  geom_errorbar(aes(ymin = Estimate - 1.96 * `Std. Error`,
                    ymax = Estimate + 1.96 * `Std. Error`),
                width = 0.2, color = "grey35") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  scale_fill_manual(values = c("Positive" = "#1D9E75", "Negative" = "#E24B4A")) +
  labs(title    = "Standardised Coefficients — Predictors of Sleep Quality",
       subtitle = "95% CI bars; longer bar = stronger independent effect",
       x = NULL, y = "Standardised Coefficient (β)",
       caption = "Fig 11") +
  theme_sleep() + theme(legend.position = "none")
```

**Expected output:** A horizontal bar chart where stress level has the longest negative bar, followed by BMI and heart rate. Physical activity has a short positive bar; daily steps and age are near zero. This directly answers which factors matter most.

------------------------------------------------------------------------

# 4 Saving Outputs

```{r save-plots, eval=FALSE}
# Save all key plots
plots_to_save <- list(
  fig08_occ_gender   = last_plot(),   # replace with named objects as needed
  fig11_factor_rank  = last_plot()
)

for (nm in names(plots_to_save)) {
  ggsave(paste0(nm, ".png"), plots_to_save[[nm]],
         width = 7, height = 5, dpi = 180)
}
```

------------------------------------------------------------------------

# 5 Summary of Findings and Conclusions

------------------------------------------------------------------------

This project examined 374 individuals from the Sleep Health and Lifestyle Dataset to determine which factors most strongly influence sleep quality and the presence of sleep disorders (insomnia, sleep apnea, or none).

**Stress is the dominant modifiable predictor.** As stress increased from level 3 to 8, average sleep quality fell from 8.97 to 5.86 — a 35% decline — and sleep duration dropped from 8.2 to 6.1 hours. The Kruskal-Wallis test confirmed that stress differences across disorder groups far exceed chance (p = 3.8×10⁻¹⁴). In the multiple regression, stress carried the largest standardised coefficient of all predictors, explaining the greatest share of variance in sleep quality independently of other factors.

**BMI acts as a near-binary switch for disorder risk.** Among normal-BMI individuals, 93.8% were disorder-free; among overweight individuals only 12.8% were disorder-free; and 100% of obese individuals had a disorder. The chi-square test (χ² = 228.5, df = 6, p \< 2.2×10⁻¹⁶) confirmed that the BMI–disorder association is not due to chance. In the multinomial logistic model, BMI had large positive coefficients for both disorder types, positioning it as a structural, threshold-like risk factor rather than a continuous gradual one.

**Insomnia and sleep apnea have distinct driver profiles.** The insomnia group showed the highest stress (mean 5.87) and the lowest physical activity (mean 46.8 min/day), pointing to a stress-and-inactivity driver. In contrast, the sleep apnea group paradoxically had the highest activity (74.8 min/day) yet still had near-universal disorder presence, consistent with sleep apnea being driven primarily by BMI-related airway physiology rather than behavioural factors.

**Heart rate is a secondary significant predictor.** With a standardised coefficient of approximately −0.6 in the MLR, higher resting heart rate is independently associated with lower sleep quality, even after controlling for stress and BMI. This may reflect the physiological pathway through which stress and excess body weight elevate sympathetic nervous system activity.

**Physical activity, daily steps, and age show weaker independent effects.** While activity level differs across disorder groups at the bivariate level, its partial regression coefficient becomes non-significant once stress and BMI are held constant, suggesting that the activity–sleep link is partly mediated by these two stronger predictors.

**Occupation and gender add nuance.** High-stress occupations (nursing, sales) are associated with lower sleep quality across both genders, whereas sedentary professional roles (engineering, accounting) tend to show better outcomes — a direction worth pursuing in further analysis with a larger, stratified dataset.

**Overall conclusion:** A two-factor risk profile — *high stress + excess BMI* — captures the majority of disorder risk in this dataset. Interventions targeting stress reduction (for insomnia) and weight management (for sleep apnea) are most likely to yield measurable improvements in population sleep health. Future work should include longitudinal data, objective actigraphy measurements, and clinical covariates (medication use, comorbidities) to strengthen causal inference.

------------------------------------------------------------------------

*Report prepared by Team Synapse. Dataset: Kaggle — Sleep Cycle and Productivity.*
